---
title: "Visualization: Bar Charts"
author: Zach del Rosario
date: 2020-05-13
output: github_document
time: 30
reading: 30
---

*Purpose*: *Bar charts* are a key tool for EDA. In this exercise, we'll learn
how to construct a variety of different bar charts, as well as when---and when
*not*---to use various charts.

*Reading*: [Bar Charts](https://rstudio.cloud/learn/primers/3.2)
*Topics*: (All topics)
*Reading Time*: ~30 minutes

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

__q1__ In the reading, you learned the relation between `geom_bar()` and
`geom_col()`. Use that knowledge to convert the following `geom_bar()` plot into
the same visual using `geom_col()`.

```{r q1-task}
#original code
mpg %>%
  ggplot(aes(trans)) +
  geom_bar()

#altered code
mpg %>%
  count(trans) %>%
  ggplot() +
  geom_col(mapping = aes(x = trans, y = n))
```

The reading mentioned that when using `geom_col()` our x-y data should be
`1-to-1`. This next exercise will probe what happens when our data are not
`1-to-1`, and yet we use a `geom_col()`. Note that a
[one-to-one](https://en.wikipedia.org/wiki/Injective_function) function is one
where each input leads to a single output. For the `mpg` dataset, we can see
that the pairs `cty, hwy` clearly don't have this one-to-one property:

```{r mpg-not-1-to-1}
## NOTE: Run this chunk for an illustration
mpg %>% filter(cty == 20)
```

__q2__ The following code attempts to visualize `cty, hwy` from `mpg` using
`geom_col()`. There's something fishy about the `hwy` values; what's wrong here?

*Hint*: Try changing the `position` parameter for `geom_col()`.

```{r q2-task}
#original
mpg %>%
  ggplot(aes(x = cty, y = hwy)) +
  geom_col()

#modified
mpg %>%
  ggplot(aes(x = cty, y = hwy)) +
  geom_col(position = 'dodge')
```

**Observations**:
- with the bars stacked on top of each other, the hwy gas mileage of the most prevalent cars looks way too high. When we split them with 'dodge' then we get to see what is actually going on at each fuel mileage.

A more standard way to visualize this kind of data is a *scatterplot*, which
we'll study later. For now, here's an example of a more effective way to
visualize `cty` vs `hwy`:

```{r q2-compare}
## NOTE: Run this chunk for an illustration
mpg %>%
  ggplot(aes(cty, hwy)) +
  geom_point()
```

__q3__ The following are two different visualizations of the `mpg` dataset.
Document your observations between the `v1` and `v2` visuals. Then, determine
which---`v1` or `v2`---enabled you to make more observations. What was the
difference between the two visuals?

```{r q4-v1}
## TODO: Run this code without changing, describe your observations on the data
mpg %>%
  ggplot(aes(class, fill = class)) +
  geom_bar()
```

**Observations**:
- SUVs are most prevalent
- 2-seaters are least prevalent and are salmon-colored

```{r q4-v2}
## TODO: Run this code without changing, describe your observations on the data
mpg %>%
  ggplot(aes(class, fill = drv)) +
  geom_bar()
```

**Observations**:
- Not only are SUVs most common but they are mostly 4wd
- All pickup trucks are 4wd
- All minivans are fwd
- All 2 seaters are rwd.

**Compare `v1` and `v2`**:

- Which visualization---`v1` or `v2`---enabled you to make more observations?
- What was the difference between `v1` and `v2`?
- There was an additional dimension of data plotted on v2 relative to v1. They're both pretty, but v2 is more informative. 

__q4__ The following code has a bug; it does not do what its author intended.
Identify and fix the bug. What does the resulting graph tell you about the
relation between `manufacturer` and `class` of cars in this dataset?

*Note*: I use a `theme()` call to rotate the x-axis labels. We'll learn how to
do this in a future exercise.

```{r q4-task}
#original
mpg %>%
  ggplot(aes(x = manufacturer, fill = class), position = "dodge") +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 270, vjust = 0.5, hjust = 0))

#modified
mpg %>%
  ggplot(aes(x = manufacturer, fill = class)) +
  geom_bar(position = 'dodge') +
  theme(axis.text.x = element_text(angle = 270, vjust = 0.5, hjust = 0))
```

**Observations**
- Not only is "Dodge" an unfortunate name collision, but it was in the ggplot call instead of the geom_bar call.
- This does not actually set the position correct and doesn't seem to throw an error either.

__q5__ The following graph is hard to read. What other form of faceting would
make the visual more convenient to read? Modify the code below with your
suggested improvement.

```{r q5-task}
mpg %>%
  ggplot(aes(x = cyl)) +
  geom_bar() +
  facet_wrap(~ manufacturer)
```
facet_wrap
<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis01-bar-charts-assignment.Rmd).
