---
title: "RMS Titanic"
author: "Elliott Donlon"
date: 15 July 2020
output:
  github_document:
    toc: true
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

__q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

- Class
- Sex
- Age
- Survived
- n

__q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of passengers. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- The wikipedia article states that 2224 passengers and crew were aboard. This difference is 23. This relatively small difference (~1%) could be attributed to the fact that some passengers that bought tickets didn't show for various reasons while others left part-way through the journey.

__q3__ Create a plot showing the count of passengers who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = n, fill = Sex))
  #facet_wrap( ~ sex)
```

**Observations**:

- We can see that there are more male crew members that survived the incident than females
- More females survived in 1st and 2nd class
- About equal proportions survived in 3rd class
- There are less second-class people that survived of any class
- How are these data skewed by the number of people in each class to begin with?

# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

__q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3.

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = Prop, fill = Sex))
  #facet_wrap( ~ sex)
```

**Observations**:

- It seems that the prop variable is totaled by each combination of class, age and sex so when viewed like this, the prop value is kind of meaningless. 
- If we re-compute the totals we can get more accurate proportions at this depth of analysis (see figure q4-task_b)
- If we split the data like in the q5 plot, we get a better idea of what is going on proportionally.

```{r q4-task_b}
df_prop_class <-
  df_titanic %>%
  group_by(Class, Sex) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop_class

df_prop_class %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = Prop, fill = Sex), position = 'dodge')
```

**Observations**:

- Here, since we changed the way we calculate the proportions by grouping, we can clearly see what proportion of sexes in each class are survivors. 
- We can clearly see that females had much higher survival rate than their male counterparts in each class.
- 3rd class was still lowest-survival at about 37% survival for females and about 13% for males.
- Almost all females in 1st class survived and about 87% of female crw members survived while 75% of 2nd class females survived.

__q5__ Create a plot showing the group-proportion of passengers who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = Prop, fill = Sex), position = 'dodge') +
  facet_wrap( ~ Age)
```

**Observations**:

- Children had a much higher survival rate than their adult counterparts in 1st and 2nd class.
- Children in 3rd class were about equal in survival rate to their adult counterparts. 
- (Thank goodness there were no child crew members)
- All children in 1st and 2nd class survived (yay)!
- More female children survived than male children where applicable (3rd class).
- Hypothesis: "Women and children first" *could* be a mindset that hinders the survival of able-bodied men in this catastrophe. If we assume that is true, which men are most "chivalrous"?

```{r q5-task_chivalry}
df_prop_chivalry <-
  df_titanic %>%
  filter(Age == "Adult") %>%
  group_by(Class, Sex) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  filter(Survived == "Yes") %>%
  select(-n, -Total)%>%
  pivot_wider(
   names_from = Sex,
   values_from = Prop
  ) %>%
  mutate(chivalry = Female / Male)
df_prop_chivalry
```

**Summary Observations**

- Overall, more females survived in each class
- Children in 1st and 2nd class were most likely to survive of all groups
- From the previous table we can see that for this definition of the "chivalry" metric, the men in 2nd class were 4 to 5x more "chivalrous" than the other classes.

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
